### 2022-11-11 NSF2022/Description/specificaim2.tex
working on SA2

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. The success
of Specific Aims 1 and 3 depends on the ability to accurately and
efficiently simulate dense suspensions of granules governed by
equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance}. Because of their
ability to resolve complex unbounded domains and achieve high-order
accuracy, integral equation methods will be a central theme of the
computational tools. As discussed in Section~\ref{sec:bie}, numerical
methods to solve IEs have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, and
elements of these formulations have been developed in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. In our current
implementation, we can simulate $\mathcal{O}(100)$ two-dimensional
granules, and the proposed research will allow for the suspensions with
100X--1000X more two-dimensional granules and $\mathcal{O}(1000)$
three-dimensional granules.

@@

The order parameter is governed by the PDE~\eqref{eqn:phase} in the
solvent phase $\Omega(t)$. For the single-well $W(\phi) = \phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have demonstrated that
IEs are a powerful tool to solve for the order
parameter~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}. However,
for a general double-well potential $W(\phi)$, the resulting PDE is
nonlinear, motivating the following problem:

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 3.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$W(\phi) = \phi^2$---this choice results in the linear screened Laplace
equation. For a general double well potential $W$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution as the root
finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (dF^N)^{-1} F(\phi^N)$, where $dF^N =
-\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v = F[\phi^N]$, which
is challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $dF^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$. We
test the efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $dF$
with a second-order finite difference stencil. The double well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

Having demonstrated that the inexact Newton method converges in one
dimension, it will be coupled with IE formulations to
solve~\eqref{eqn:phase} in two and three dimensions. The solution of
$-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$ can be written
as $v = v^P + v^H$ where $v^P$ is a volume potential and $v^H$ is a
homogeneous solution of the PDE, which is represented as a layer
potential, and is used to satisfy the boundary condition. The volume
potential is
\begin{align*}
  v^{P}(\xx) = \int_{\Omega(t)} G(\xx,\yy) F[\phi^N](\yy)\, d\yy, 
    \quad \xx \in \Omega(t),
\end{align*}
where $G$ is the fundamental solution of $-\epsilon^2 \Delta +
\tfrac{\beta}{2}$. Traditionally, volume potentials are more challenging
to compute than layer potentials, but new methods have recently been
developed that allows them to be computed accurately and
efficiently~\cite{fry-leh-tor2018, and-zhu-vee2022}. This approach of
combining volume and layer potentials has successfully been used by PI
BQ~\cite{kro-qua2010} and others~\cite{fry-kro-tor2019, kli-ask-kro2020,
che-hua-lei2005}.
% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 4.} How can novel quadrature, time adaptivity, fast
  summation, and collision-detection be combined to accurately and
  efficiently solve the governing
  equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance} in bounded and
  unbounded domains in two and three dimensions?
\end{quotation}

\begin{wrapfigure}[]{l}{2.1in}
  \centering
  \includegraphics[width=2in]{figures/SA2Figures/confinement.pdf}
  \caption{\label{fig:stenosis} A granule-based vesicle passing through
  a constricted geometry.}
\end{wrapfigure}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}. The strong confinement introduces an aditional
computational challenge---acuartely computing integrals between
nearly-touching bodies. This issue is well-known in the IE community and
several algorithms have been developed in the last 30 years, including
panel-based methods~\cite{helsing_2008a, helsing_tutorial_2012},
Quadrature by Expansions (QBX)~\cite{din-hua-mar-tan2021,
klo-bar-gre-one2013, ste-bar2022, af2018sisc}, Barycentric-type formulae
for Cauchy integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991},
and interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

\begin{wrapfigure}[]{r}{2.1in}
  \centering
  \input{figures/near_sing.tikz}
  \caption{\label{fig:interpquad} An interpolation-based quadrature
  method that works in two and three dimensions. The spectrally-accurate
  trapezoid rule is used in points in $\mathtt{Far}(\Gamma_i)$, while
  one dimensional interpolant is used for points in
  $\mathtt{Near}(\Gamma_i)$. $h$ is the arclength spacing.}
\end{wrapfigure}

@@

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional reqired
computational tool is a collision-aware adaptive time stepping method.
The order parameter results in the granules attracting one another and
this leads to physical contact in finite time. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions with by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2022}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.

>>>>

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. The success
of Specific Aims 1 and 3 depends on the ability to accurately and
efficiently simulate dense suspensions of granules governed by
equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance}. Because of their
ability to resolve complex unbounded domains and achieve high-order
accuracy, integral equations (IEs) will be a central theme of the
computational tools.

@@

The order parameter is governed by the PDE~\eqref{eqn:phase} in the
solvent phase $\Omega(t)$. For the single-well $W(\phi) = \phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have used IEs to solve
for the order parameter~\cite{Fu2018_SIAM, FuQuRyYo22,
fu-ryh-qua-you2022}. However, for a double-well potential $W(\phi)$, the
resulting PDE is nonlinear, motivating the following problem:

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 4.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

For a general double-well potential $W(\phi)$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution
of~\eqref{eqn:phase} as a root finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (dF^N)^{-1} F(\phi^N)$, where $dF^N =
-\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v = F[\phi^N]$, which
is challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double-well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $dF^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$. We
test the efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $dF$
with a second-order finite difference stencil. The double well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

Having demonstrated that the inexact Newton method converges in one
dimension, it will be coupled with IE formulations to
solve~\eqref{eqn:phaseRoot} in two and three dimensions. The solution of
$-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$ can be written
as $v = v^P + v^H$ where $v^P$ is a volume potential and $v^H$ is a
homogeneous solution of the PDE, which is represented as a layer
potential, and is used to satisfy the boundary condition. The volume
potential is
\begin{align*}
  v^{P}(\xx) = \int_{\Omega(t)} G(\xx,\yy) F[\phi^N](\yy)\, d\yy, 
    \quad \xx \in \Omega(t),
\end{align*}
where $G$ is the fundamental solution of $-\epsilon^2 \Delta +
\tfrac{\beta}{2}$. Traditionally, volume potentials are more challenging
to compute than layer potentials, but new methods have recently been
developed that allow them to be computed accurately and
efficiently~\cite{fry-leh-tor2018, and-zhu-vee2022}. This approach of
combining volume and layer potentials has successfully been used by PI
Quaife~\cite{kro-qua2010} and others~\cite{fry-kro-tor2019,
kli-ask-kro2020, che-hua-lei2005} in other contexts, such as time
stepping.

As discussed in Section~\ref{sec:bie},
numerical methods to solve IEs have several advantages over their
PDE-based counterpart. The PIs have demonstrated that IEs are a powerful
tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, and
elements of these formulations have been developed in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. In our current
implementation, we simulate $\mathcal{O}(100)$ two-dimensional granules,
and the proposed research will allow for the suspensions with
100X--1000X more two-dimensional granules and $\mathcal{O}(1000)$
three-dimensional granules.

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 5.} How can novel quadrature, time adaptivity, fast
  summation, and collision-detection be combined to accurately and
  efficiently solve the governing
  equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance} in bounded and
  unbounded domains in two and three dimensions?
\end{quotation}

\begin{wrapfigure}[]{l}{2.1in}
  \centering
  \includegraphics[width=2in]{figures/SA2Figures/confinement.pdf}
  \caption{\label{fig:stenosis} \footnotesize A granule vesicle passing
  through a constricted geometry.}
\end{wrapfigure}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}.

The confinement introduces an additional computational
challenge---accurately computing integrals between nearly-touching
bodies. This issue is well-known in the IE community and several
algorithms have been developed in the last 30 years, including
panel-based methods~\cite{helsing_2008a, helsing_tutorial_2012},
Quadrature by Expansions (QBX)~\cite{din-hua-mar-tan2021,
klo-bar-gre-one2013, ste-bar2022, af2018sisc}, Barycentric-type formulae
for Cauchy integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991},
and interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

\begin{wrapfigure}[]{r}{2.1in}
  \centering
  \input{figures/near_sing.tikz}
  \caption{\label{fig:interpquad} \footnotesize An interpolation-based
  quadrature method that works in two and three dimensions. The
  spectrally-accurate trapezoid rule is used in points in
  $\mathtt{Far}(\Gamma_i)$, while one dimensional interpolant is used
  for points in $\mathtt{Near}(\Gamma_i)$. $h$ is the arclength
  spacing.}
\end{wrapfigure}

@@

While the inclusion of high-order quadrature and fast methods is
necessary to simulate dense suspensions, an additional reqired
computational tool is a collision-aware adaptive time stepping method.
The order parameter results in the granules attracting one another and
this leads to physical contact in finite time. Our current
implementation uses a repulsive Lennard-Jones force, but this introduces
stiffness, thereby requiring small time steps, and this can be a
bottleneck when simulating dense suspensions. Therefore, we will develop
a high-order collision-avoiding adaptive time stepping method. This
method will avoid collisions with by using the minimal separation
distance to impose a non-overlapping constraint. This results in a
linear complementarity problem that can be solve efficiently with a
projected gradient descent method~\cite{Yan2019}. High-order time
adaptivity will be based on PI Quaife's previous work on
suspensions~\cite{qua-bir2016}, and this method has proven to be a
robust method to simulate dense suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017, qua-gan-you2021}. Following this previous work, the
proposed work will use a spectral deferred correction method since it
iteratively applies a low-order, single-step method to achieve
high-order accuracy. At each time step, we will compute the total force
and torque of the system which is computationally cheap and physically
zero, and therefore provides an estimate of the local truncation error.
PI Quaife's experience is that error estimates based on physical
constraints, such as force- and torque-free conditions, appropriately
determine adaptive time step sizes so that the dynamics are resolved
without the computational expense of techniques such as embedded
Runge-Kutta, step-doubling, and Richardson extrapolation.
